Transferability of clinical prediction models for early trauma care

### 2019-06-04 ManuscriptMarkdown.Rmd
Add files via upload

---
title: "Transferability of clinical prediction models for early trauma care"
author: "Martin Henriksson"
date: "190602"
output: html_document
bibliography: library.bib
csl: bmcemerg.csl
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
## LOAD FUNCTIONS, ENVIROMENT AND TABLES
  ## Source functions
  source("functions/WorkTableCreator.R")
  ## Original results
  results <- readRDS("output/original.results.Rds")
  ## Compiled results with UI
  result.tables <- readRDS("result.tables.Rds")
  ## Compiled results with UI as data frame
  result.tables.data.frame <- readRDS("result.tables.data.frame.Rds")
  ## Create table for ease of data retrival
  results.all.data <- WorkTableCreator(result.tables.data.frame)
  ## Sample characteristics table
  # table.one <-

@@

# Introduction
Trauma results in millions of deaths annually across the globe, and road traffic injuries are among the top 10 leading causes of death worldwide [@RN1; @RN21]. Each year 9% of the world’s deaths are because of trauma, and predictions indicate that this number is likely to increase [@RN42].

@@

Systems to determine the level of trauma can be based on clinical prediction models. Such models are continuously being developed and studied, examining different parameters and different contexts. Models differ in quality and characteristics, but generally perform well predicting survival (6). Many of these models are developed in a single standardized setting such as a large metropolitan area or a major trauma centre but are then implemented in different settings, such as non-trauma centres.

What is not fully understood is how this transfer affects model performance. Previous research has shown that model performance in terms of calibration can be adversely affected (7). However, the aforementioned research studied model transfers between substantially different settings and did not assess more clinically relevant performance measures, such as misclassification.

@@

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age. It is estimated that up to 10% of all major trauma patients suffer some form of disability. The societal effects of trauma are also substantial; with the loss of work days comparable to those lost from cardiovascular and malignant diseases combined (5).

The trauma care organization in Sweden is usually clearly defined in major metropolitan areas: trauma patients are transported directly to a predesignated hospital with specific trauma competency. In large, more rural parts of the country, trauma patients are instead transported to the nearest hospital for stabilization. Once stabilized, they may then be transported to hospitals with more advanced trauma care capacity (19). Most hospitals in Sweden use some form of system to categorize the level of trauma. These systems are usually based on patient vital signs and injury mechanism, as reported by the EMS, or as registered in the emergency department (16).

Data was obtained from SweTrau between 2011 and 2016. Today 95,5% (52 of 55) of Swedish hospitals record trauma cases in SweTrau. Currently the registry includes 55 000 cases, and the board encourages its use in both academic research and more local quality improvement initiatives (20).

The SweTrau inclusion criteria are: Traumatic event with subsequent activation of hospital trauma protocol, admitted patients with NISS (New Injury Severity Score) > 15 and patients transferred to the hospital within 7 days of traumatic event with NISS > 15. SweTrau excludes patients if the only traumatic event is chronic subdural hematoma or if hospital trauma protocol is activated without traumatic event (21).

## Participants 
The eligibility criteria were patients registered in SweTrau, age 15 or above. This age was decided as the study aims to study adult trauma and not paediatric trauma which differs in physiology, triage and initial care (22). The Swedish guidelines for trauma activation define children as age <15 (16). Also, several guidelines and protocols listed in ATLS use age ≥15 as cut-off (9). Patient age was obtained from the SweTrau register.

## Variables
### Model predictors
The clinical prediction models included the predictors SBP, RR and GCS on arrival to hospital. The rationale for including these three predictors is that they are part of many established clinical prediction models for early trauma care, such as the Revised Trauma Score (23).

@@

### Study outcome
ISS > 15 was used as the gold standard to define trauma severity as major trauma, and hence patients with ISS ≤ 15 were considered minor trauma (24). Overtriage was defined as the event when a clinical prediction model classified a patient with ISS ≤ 15 as major trauma, and undertriage as the event when a clinical prediction model classified a patient with ISS > 15 as minor trauma. Overtriage rate was defined as the number of overtriaged patients divided by all patients. Undertriage rate was defined as the number of undertriaged patients divided by all patients. Mistriage rate was defined as the sum of the over- and undertriage rates.

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into two subsamples using a temporal split based on the date of traumatic event. The earlier subsample was the development sample, and the later subsample the validation sample. The development sample contained 70 events (events being patients who died within 30 days of the trauma) and all non-events (non-events being patient survival 30 days past the trauma) during the same time. The rationale for including 70 events is that we needed at least 10 events per free parameter in the logistic regression to obtain stable coefficient estimates (25). The validation sample instead contained 100 events and at least 100 non-events, which was suggested as the minimum number by Vergouwe in 2005 for external validation samples (26). See figure 1 for example using the high- and low volume centre data set, with high volume centres being those in the top quartile of number of patients registered.

@@

### Sequence of analysis
The programming language R was used for all analyses (27), all progress was uploaded to a GitHub repository (28). Analyses were performed in the sequence of model development, model validation and finally model comparison. These steps were repeated in each data set. Below we used the transfer of a model from a high-volume sample to a low volume sample as an example to describe the complete procedure.

### Model development
In the model development step a clinical prediction model was developed in the high-volume centre development sample. The model was developed using logistic regression as implemented in the R function glm. The dependent variable was all cause mortality within 30 days of trauma and independent variables SBP, RR, and GCS modelled as previously described. To avoid overfitting the model we used a bootstrap procedure to estimate a linear shrinkage factor that we then applied to the model coefficients (29). The shrunk model was then used to estimate the probability of all cause 30-day mortality in the development sample. A gridsearch was performed across estimated probabilities in the development sample to identify the cut-off that optimised overtriage keeping undertriage at less than 5% (11). This cut-off was then used to classify patients as major or minor trauma.

@@

### Missing data
We used multiple imputation using chained equations, as implemented in the R package mice, to handle missing data (30). The number of imputations created for each data set was equal to the highest percentage of missing data in that data set. Quantitative variables were imputed using predictive mean matching and qualitative variables were imputed using logistic regression. SBP and RR was transformed as restricted cubic splines before imputation and imputed as just another variable. All analyses outlined above were then conducted separately in each imputed dataset. We present the combined results as the median point estimate across imputations along with an empirical bootstrap of the 25th and 75th percentiles across imputations, i.e. the lower bound of the presented interval is the lower bound of a 95% CI of the 25th percentile and the upper bound is the upper bound of a 95% CI of the 75th percentile. This combined CI was referred to as an Uncertainty Interval (UI) and was used to express the added uncertainty associated with the imputation procedure to handle missing data, as such it is more conservative than a standard 95% CI.

@@

## Comparison
Model performance after transfer was determined for each model being transferred to the other sample in the data set, in all data sets (Table 3). When transferred, the model with the lowest mistriage rate was the `r colnames(lowest.tran.mistriage)` model with a median mistriage rate of `r lowest.tran.mistriage`. The model with the highest mistriage rate after transfer was the `r colnames(highest.tran.mistriage)` with a median mistriage rate of `r highest.tran.mistriage`

@@

The transfer of clinical prediction models in trauma care has not previously been extensively studied. In 2016, Gerdin et al. found model transfer to adversely affect model performance in terms of calibration, but that this could be improved by updating the model (18). This study did not examine model performance in terms of calibration, neither was the effects of updating the model explored.

External validation as the evaluation of a clinical prediction model in a setting in which it was not originally developed has been studied more extensively. Studies using simulated clinical prediction models based on different predictors showed a decline in model performance when being externally validated (33, 34). A study from 2018 examining the external validity of prediction models for coronary artery disease (CAD) (35), showed models significantly underestimating the probability of CAD. This study did not find the same dramatic decrease in model performance after transfer. However, the combined results of this study, the studies on model transfer and the studies on external validation warrants further caution when transferring models between different settings.

@@

The predetermined analysis plan was to be followed in a step-by-step fashion, to avoiding confirmation bias. Programming challenges and late-time code inclusions required a flexible approach to this analysis plan, sometimes with revisions of previous steps. This is not optimal, and efforts were made to not deviate in any major way. Using GitHub, all code was made public for inspection and to ensure reproducibility (28).

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Gerdin Wärnberg
- Martin Henriksson
- Johanna Berg
- Dell Saulnier
date: "June 04, 2019"
documentclass: "article"
output: html_document
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
## LOAD FUNCTIONS, ENVIROMENT AND TABLES
  ## Source functions
  source("functions/WorkTableCreator.R")
  ## Original results
  results <- readRDS("output/original.results.Rds")
  ## Compiled results with UI
  result.tables <- readRDS("result.tables.Rds")
  ## Compiled results with UI as data frame
  result.tables.data.frame <- readRDS("result.tables.data.frame.Rds")
  ## Create table for ease of data retrival
  results.all.data <- WorkTableCreator(result.tables.data.frame)
  ## Sample characteristics table
  # table.one <-

@@

---
abstract: |
  **Background:** Trauma results in millions of deaths annually. Trauma care involves many levels of the health care system, and clinical prediction models are often used to triage patients. Previous research has indicated transferring these models might adversely affect model performance.

  **Methods:** Patients age 15 or above registered between the years 2011 and 2016 in the SweTrau registry were included. Three sets of data were created: High and low volume centres, metropolitan and non-metropolitan centres, multi and single centre data. A logistic regression model was developed. Model performance was evaluated in all data sets using mistriage rate, undertriage rate and overtriage rate. Multiple imputation using chained equation was used to handle missing data. Model performance was reported as medians with 95% uncertainty interval across imputed data sets.

  **Results:** `r total.number.of.patients ` patients were included. Model validation median mistriage rate with 95% UI ranged from `r lowest.val.mistriage` to `r highest.val.mistriage`. and transferred model mistriage rate ranged from `r lowest.tran.mistriage` to `r highest.tran.mistriage` across all samples. Worst performance after transfer compared to receiving sample validation performance was found in the `r colnames(highest.comp.mistriage)` sample with a median mistriage difference of `r highest.comp.mistriage`.

  **Conclusions:** Model transfer from the `r colnames(highest.comp.mistriage)` to the other sample in the data set lead to an increased mistriage rate. Mainly due to a marked increase in overtriage. In all samples, model transfer was associated with high levels of uncertainty.

  **Keywords:** Trauma, Clinical prediction model, Transferability, Mistriage, Undertriage, Overtriage
---

# Introduction
Trauma results in millions of deaths annually across the globe, and road traffic injuries are among the top 10 leading causes of death worldwide [@RN1; @RN21]. Each year 9% of the world’s deaths are because of trauma, and predictions indicate that this number is likely to increase [@RN42].

@@

Systems to determine the level of trauma can be based on clinical prediction models. Such models are continuously being developed and studied, examining different parameters and different contexts. Models differ in quality and characteristics, but generally perform well predicting survival [@RN45]. Many of these models are developed in a single standardized setting such as a large metropolitan area or a major trauma centre but are then implemented in different settings, such as non-trauma centres.

What is not fully understood is how this transfer affects model performance. Previous research has shown that model performance in terms of calibration can be adversely affected [@RN27]. However, the aforementioned research studied model transfers between substantially different settings and did not assess more clinically relevant performance measures, such as misclassification.

@@

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age. It is estimated that up to 10% of all major trauma patients suffer some form of disability. The societal effects of trauma are also substantial; with the loss of work days comparable to those lost from cardiovascular and malignant diseases combined [@RN23].

The trauma care organization in Sweden is usually clearly defined in major metropolitan areas: trauma patients are transported directly to a predesignated hospital with specific trauma competency. In large, more rural parts of the country, trauma patients are instead transported to the nearest hospital for stabilization. Once stabilized, they may then be transported to hospitals with more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some form of system to categorize the level of trauma. These systems are usually based on patient vital signs and injury mechanism, as reported by the EMS, or as registered in the emergency department [@RN90].

Data was obtained from SweTrau between 2011 and 2016. Today 95,5% (52 of 55) of Swedish hospitals record trauma cases in SweTrau. Currently the registry includes 55 000 cases, and the board encourages its use in both academic research and more local quality improvement initiatives [@RN28].

The SweTrau inclusion criteria are: Traumatic event with subsequent activation of hospital trauma protocol, admitted patients with NISS (New Injury Severity Score) > 15 and patients transferred to the hospital within 7 days of traumatic event with NISS > 15. SweTrau excludes patients if the only traumatic event is chronic subdural hematoma or if hospital trauma protocol is activated without traumatic event [@RN29].

## Participants 
The eligibility criteria were patients registered in SweTrau, age 15 or above. This age was decided as the study aims to study adult trauma and not paediatric trauma which differs in physiology, triage and initial care [@RN46]. The Swedish guidelines for trauma activation define children as age <15 [@RN90]. Also, several guidelines and protocols listed in ATLS use age ≥15 as cut-off [@RN43]. Patient age was obtained from the SweTrau register.

## Variables
### Model predictors
The clinical prediction models included the predictors SBP, RR and GCS on arrival to hospital. The rationale for including these three predictors is that they are part of many established clinical prediction models for early trauma care, such as the Revised Trauma Score [@RN35].

@@

### Study outcome
ISS > 15 was used as the gold standard to define trauma severity as major trauma, and hence patients with ISS ≤ 15 were considered minor trauma [@RN36]. Overtriage was defined as the event when a clinical prediction model classified a patient with ISS ≤ 15 as major trauma, and undertriage as the event when a clinical prediction model classified a patient with ISS > 15 as minor trauma. Overtriage rate was defined as the number of overtriaged patients divided by all patients. Undertriage rate was defined as the number of undertriaged patients divided by all patients. Mistriage rate was defined as the sum of the over- and undertriage rates.

@@

### Development and validation sample
Each data set included at least two samples. The samples were then split into two subsamples using a temporal split based on the date of traumatic event. The earlier subsample was the development sample, and the later subsample the validation sample. The development sample contained 70 events (events being patients who died within 30 days of the trauma) and all non-events (non-events being patient survival 30 days past the trauma) during the same time. The rationale for including 70 events is that we needed at least 10 events per free parameter in the logistic regression to obtain stable coefficient estimates [@RN30]. The validation sample instead contained 100 events and at least 100 non-events, which was suggested as the minimum number by Vergouwe in 2005 for external validation samples [@RN37]. See figure 1 for example using the high- and low volume centre data set, with high volume centres being those in the top quartile of number of patients registered.

@@

### Sequence of analysis
The programming language R was used for all analyses [@RN31], all progress was uploaded to a GitHub repository [@RN59]. Analyses were performed in the sequence of model development, model validation and finally model comparison. These steps were repeated in each data set. Below we used the transfer of a model from a high-volume sample to a low volume sample as an example to describe the complete procedure.

### Model development
In the model development step a clinical prediction model was developed in the high-volume centre development sample. The model was developed using logistic regression as implemented in the R function glm. The dependent variable was all cause mortality within 30 days of trauma and independent variables SBP, RR, and GCS modelled as previously described. To avoid overfitting the model we used a bootstrap procedure to estimate a linear shrinkage factor that we then applied to the model coefficients [@RN61]. The shrunk model was then used to estimate the probability of all cause 30-day mortality in the development sample. A gridsearch was performed across estimated probabilities in the development sample to identify the cut-off that optimised overtriage keeping undertriage at less than 5% [@RN39]. This cut-off was then used to classify patients as major or minor trauma.

@@

### Missing data
We used multiple imputation using chained equations, as implemented in the R package mice, to handle missing data [@RN41]. The number of imputations created for each data set was equal to the highest percentage of missing data in that data set. Quantitative variables were imputed using predictive mean matching and qualitative variables were imputed using logistic regression. SBP and RR was transformed as restricted cubic splines before imputation and imputed as just another variable. All analyses outlined above were then conducted separately in each imputed dataset. We present the combined results as the median point estimate across imputations along with an empirical bootstrap of the 25th and 75th percentiles across imputations, i.e. the lower bound of the presented interval is the lower bound of a 95% CI of the 25th percentile and the upper bound is the upper bound of a 95% CI of the 75th percentile. This combined CI was referred to as an Uncertainty Interval (UI) and was used to express the added uncertainty associated with the imputation procedure to handle missing data, as such it is more conservative than a standard 95% CI.

@@

## Comparison
Model performance after transfer was determined for each model being transferred to the other sample in the data set, in all data sets (Table 3). When transferred, the model with the lowest mistriage rate was the `r colnames(lowest.tran.mistriage)` model with a median mistriage rate of `r lowest.tran.mistriage`. The model with the highest mistriage rate after transfer was the `r colnames(highest.tran.mistriage)` with a median mistriage rate of `r highest.tran.mistriage`.

@@

The transfer of clinical prediction models in trauma care has not previously been extensively studied. In 2016, Gerdin et al. found model transfer to adversely affect model performance in terms of calibration, but that this could be improved by updating the model [@RN27]. This study did not examine model performance in terms of calibration, neither was the effects of updating the model explored.

External validation as the evaluation of a clinical prediction model in a setting in which it was not originally developed has been studied more extensively. Studies using simulated clinical prediction models based on different predictors showed a decline in model performance when being externally validated [@RN60; @RN61]. A study from 2018 examining the external validity of prediction models for coronary artery disease (CAD) [@RN62], showed models significantly underestimating the probability of CAD. This study did not find the same dramatic decrease in model performance after transfer. However, the combined results of this study, the studies on model transfer and the studies on external validation warrants further caution when transferring models between different settings.

@@

The predetermined analysis plan was to be followed in a step-by-step fashion, to avoiding confirmation bias. Programming challenges and late-time code inclusions required a flexible approach to this analysis plan, sometimes with revisions of previous steps. This is not optimal, and efforts were made to not deviate in any major way. Using GitHub, all code was made public for inspection and to ensure reproducibility [@RN59].
